Wigglyhedra

### 2025-01-09 draft/wigglyhedra.tex
typographic changes

\begin{definition}
The \defn{zigzag algebra} \(A_m\) corresponding to the quiver \(Q_m\) depicted in~\cref{fig:am-quiver} is the quotient of \(kQ_m\) by all paths of length at least \(3\), together with the relations \((i-1, i, i+1) = 0 = (i+1, i, i-1)\) and \((i, i-1, i) = (i, i+1, i)\) for all~\(1 < i < m\).
\end{definition}

@@

\begin{proposition}
\label{prop:cohomology-pieces}
Let \(\alpha\) be any curve, and let \(X = X^{\bullet}\) be a chain complex that corresponds to \(\alpha\) under the Khovanov--Seidel reconstruction.
Recall that for any integer \(j\), the cohomology object \(H^j(X)\) is an object of \(\mathcal{A}_m\), and hence a direct sum of indecomposable objects of \(\mathcal{A}_m\).
\begin{enumerate}
\item The object \(X\) is a (spherical) object of \(\mathcal{A}_m\) up to shift if and only if \(\alpha\) is a wiggly arc. In this case, \(H^j(X) \neq 0\) for exactly one \(j\), such that \(X \cong H^j(X)[-j]\).
\item If \(\alpha\) is a general curve, each indecomposable summand of \(H^j(X)\) for any \(j\) is a spherical object of \(\mathcal{A}_m\) that arises from a wiggly arc, up to shift.
\end{enumerate}
\end{proposition}

@@

  We recall that the procedure to build up \(X\) from \(\alpha\) explicitly describes the differential.
  We analyse the possible non-zero components of the differential, checking how each interacts with the level.
  The only two possibilities are as follows:
  \begin{enumerate}
  \item a map \(P_i\langle g \rangle[h] \to P_{i\pm1}\langle g+1 \rangle[h-1]\), namely multiplication on the right by the path \((i,i\pm1)\), which does not change the level; or
  \item a map \(P_i\langle g \rangle[h] \to P_{i}\langle g+2 \rangle[h-1]\), namely multiplication on the right by the ``loop'' path \((i,i\pm1, i)\), which raises the level by one.
  \end{enumerate}
  The first possibility appears each time the curve wiggles either above the \(i\)th point or below the \((i-1)\)th point.
  The second possibility appears each time the curve makes a U-turn around the right of \(i\)th or the left of the \((i-1)\)th point.

@@

  More generally, consider the indecomposable pieces of the level-\(j\) truncations of \(X\) for various~\(j\), up to shift.
  As explained above, in terms of the curve \(\alpha\), this corresponds to breaking up the curve~\(\alpha\) into \(x\)-monotone portions.
  More precisely, we begin at an endpoint \(p_0\) of \(\alpha\) and follow the curve until it either ends at a point \(p_1\), or stops being \(x\)-monotone by making a U-turn around a point \(p_1\).
  In this case, we take the portion of \(\alpha\) between \(p_0\) and \(p_1\), calling it \(\alpha_1\), and setting its endpoints to be \(p_0\) and \(p_1\) respectively.
  Note that \(p_0 \neq p_1\) because \(\alpha\) is \(x\)-monotone between \(p_0\) and \(p_1\).
  We continue with the remainder of \(\alpha\) starting at \(p_1\), greedily taking the maximal \(x\)-monotone portion each time, to obtain pieces \(\alpha_1,\alpha_2, \ldots, \alpha_k\), which are all wiggly arcs.
  These wiggly arcs correspond to complexes \(X_1, \ldots, X_k\), which are spherical objects of \(\mathcal{A}_m\), up to shift.
  It is clear from the construction of the associated complex that these objects, up to shift, are exactly the indecomposable pieces of the cohomology truncations \(H^j(X)\) for various \(j\).
  This argument proves the second part of the proposition.
\end{proof}

@@

\begin{proof}
  The proof of this proposition is an easy consequence of analysing the cohomology filtration from the point of view of curves on the disk with marked points.
  Let \(X_1, \ldots, X_k\) be the objects in \(T\) without repetition, which arise from wiggly arcs \(\alpha_1, \ldots, \alpha_k\) as explained in the proof of~\cref{prop:cohomology-pieces}.
  Let \(\beta = (i,j,A,B)\) and \(\beta' = (i',j', A',B')\) be two distinct wiggly arcs among \(\{\alpha_1, \ldots,\alpha_k\}\).
  We check that \(\beta\) and \(\beta'\) are compatible in the sense of~\cref{def:compatible}.

@@

  Now suppose for contradiction that \(T\) contains both external edges.
  Then some portion of \(\alpha\) curves around each of the extreme endpoints \(0\) and \(n+1\).
  Consider the outermost strands of \(\alpha\) that curve around \(0\) and \(n+1\) respectively.
  These must then connect to each other, resulting in \(\alpha\) having a separate closed component that winds around the outside of the \(n+2\) points.
  This is not possible because \(\alpha\) is a connected curve with two endpoints.
  Therefore \(T\) cannot contain both external edges.
\end{proof}

@@

With our path-length grading, note that a morphism in \(\Hom_{\mathcal{C}_m}(X,X'\langle -s_2 \rangle[s_1])\) has level \(s_1 - s_2\); that is, it takes an object \(X\) of level \(0\) to the object \(X'\langle -s_2 \rangle[s_1]\) of level \(s_1 - s_2\).
  We can thus obtain the dimension of all morphisms of level \(\ell\) by setting \(q_1 = q\) and \(q_2 = q^{-1}\) and taking the coefficient of \(q^{\ell}\) in the expression~\eqref{eq:bigraded-homs}.
    The dimension of \(\mathbb{E}^1(X,X')\), is therefore the coefficient of \(q\) under this specialisation.

It can be checked that under the path-length grading, the space of morphisms arising from an intersection of curves of ``crossing'' type as well as ``non pointed'' type contains exactly one morphism (up to equivalence and scalar multiplication) of level one; that is, a unique element (up to scalar multiplication) of \(\mathbb{E}^1\).
Morphisms arising from other intersections do not contribute to \(\mathbb{E}^1\).
This calculation proves the proposition.
\end{proof}

@@

\begin{proof}
  The proof is once again a consequence of analysing the extensions explicitly via the complexes from the Khovanov--Seidel reconstruction.
  We do not reproduce the proof here, as the argument is contained in the proof of~\cite[Thm.~5.5]{tho:06}.
  (See also~\cite[Fig.~4]{tho:06}.)
\end{proof}
Note that if we write down the relations in the Grothendieck group of our category that correspond to the extensions from each case of~\cref{prop:extensions}, then we obtain precisely the linear dependences that appear between the \(\b{g}\)-vectors in~\cref{lem:linearDependences}.
More precisely, we have that
\begin{itemize}
\item if \(\alpha\) and \(\alpha'\) are not pointed, then
  \([Y] = [X] + [X'] = [Y']\),
  from which we see \([X] + [X'] = ([Y] + [Y'])/2\);
\item if \(\alpha\) and \(\alpha'\) are crossing, then
  \([X] + [X'] = [Y] + [Y'].\)
\end{itemize}
This suggests that there should be a way to construct the \(\b{g}\)-vectors categorically in such a way that the \(\b{g}\)-vectors also map to the Grothendieck group of the category.
The \(\b{g}\)-vector of an object \(X\) corresponding to a wiggly arc \(\alpha\) can, of course, be reconstructed from \(\alpha\).
However, at the moment we do not have a purely categorical construction of the \(\b{g}\)-vector of \(X\) that does not refer to \(\alpha\), and it would be interesting to find this.

>>>>

\begin{definition}
The \defn{zigzag algebra} \(A_m\) corresponding to the quiver \(Q_m\) depicted in~\cref{fig:am-quiver} is the quotient of \(kQ_m\) by all paths of length at least \(3\), together with the relations \((i-1, i, i+1) = 0\), \((i+1, i, i-1)=0\) and \((i, i-1, i) = (i, i+1, i)\) for all~\(1 < i < m\).
\end{definition}

@@

\begin{proposition}
\label{prop:cohomology-pieces}
Let \(\alpha\) be any curve, and let \(X = X^{\bullet}\) be a chain complex that corresponds to \(\alpha\) under the Khovanov--Seidel reconstruction.
Recall that for any integer \(j\), the cohomology object~\(H^j(X)\) is an object of \(\mathcal{A}_m\), and hence a direct sum of indecomposable objects of \(\mathcal{A}_m\).
\begin{enumerate}
\item The object \(X\) is a (spherical) object of \(\mathcal{A}_m\) up to shift if and only if \(\alpha\) is a wiggly arc. In this case, \(H^j(X) \neq 0\) for exactly one \(j\), such that \(X \cong H^j(X)[-j]\).
\item If \(\alpha\) is a general curve, each indecomposable summand of \(H^j(X)\) for any \(j\) is a spherical object of \(\mathcal{A}_m\) that arises from a wiggly arc, up to shift.
\end{enumerate}
\end{proposition}

@@

  We recall that the procedure to build up \(X\) from \(\alpha\) explicitly describes the differential.
  We analyse the possible non-zero components of the differential, checking how each interacts with the level.
  The only two possibilities are as follows:
  \begin{enumerate}
  \item a map \(P_i\langle g \rangle[h] \to P_{i\pm1}\langle g+1 \rangle[h-1]\), namely multiplication on the right by the path~\((i,i\pm1)\), which does not change the level; or
  \item a map \(P_i\langle g \rangle[h] \to P_{i}\langle g+2 \rangle[h-1]\), namely multiplication on the right by the ``loop'' path~\((i,i\pm1, i)\), which raises the level by one.
  \end{enumerate}
  The first possibility appears each time the curve wiggles either above the \(i\)th point or below the~\((i-1)\)th point.
  The second possibility appears each time the curve makes a U-turn around the right of \(i\)th or the left of the \((i-1)\)th point.

@@

  More generally, consider the indecomposable pieces of the level-\(j\) truncations of \(X\) for various~\(j\), up to shift.
  As explained above, in terms of the curve \(\alpha\), this corresponds to breaking up the curve~\(\alpha\) into \(x\)-monotone portions.
  More precisely, we begin at an endpoint \(p_0\) of \(\alpha\) and follow the curve until it either ends at a point \(p_1\), or stops being \(x\)-monotone by making a U-turn around a point \(p_1\).
  In this case, we take the portion of \(\alpha\) between \(p_0\) and \(p_1\), calling it \(\alpha_1\), and setting its endpoints to be \(p_0\) and \(p_1\) respectively.
  Note that \(p_0 \neq p_1\) because \(\alpha\) is \(x\)-monotone between~\(p_0\) and~\(p_1\).
  We continue with the remainder of \(\alpha\) starting at \(p_1\), greedily taking the maximal \(x\)-monotone portion each time, to obtain pieces \(\alpha_1,\alpha_2, \ldots, \alpha_k\), which are all wiggly arcs.
  These wiggly arcs correspond to complexes \(X_1, \ldots, X_k\), which are spherical objects of \(\mathcal{A}_m\), up to shift.
  It is clear from the construction of the associated complex that these objects, up to shift, are exactly the indecomposable pieces of the cohomology truncations \(H^j(X)\) for various \(j\).
  This argument proves the second part of the proposition.
\end{proof}

@@

\begin{proof}
  The proof of this proposition is an easy consequence of analysing the cohomology filtration from the point of view of curves on the disk with marked points.
  Let \(X_1, \ldots, X_k\) be the objects in~\(T\) without repetition, which arise from wiggly arcs \(\alpha_1, \ldots, \alpha_k\) as explained in the proof of~\cref{prop:cohomology-pieces}.
  Let \(\beta = (i,j,A,B)\) and \(\beta' = (i',j', A',B')\) be two distinct wiggly arcs among~\(\{\alpha_1, \ldots,\alpha_k\}\).
  We check that \(\beta\) and \(\beta'\) are compatible in the sense of~\cref{def:compatible}.

@@

  Now suppose for contradiction that \(T\) contains both external edges.
  Then some portion of \(\alpha\) curves around each of the extreme endpoints \(0\) and \(n+1\).
  Consider the outermost strands of \(\alpha\) that curve around \(0\) and \(n+1\) respectively.
  These must then connect to each other, resulting in~\(\alpha\) having a separate closed component that winds around the outside of the \(n+2\) points.
  This is not possible because \(\alpha\) is a connected curve with two endpoints.
  Therefore \(T\) cannot contain both external edges.
\end{proof}

@@

With our path-length grading, note that a morphism in \(\Hom_{\mathcal{C}_m}(X,X'\langle -s_2 \rangle[s_1])\) has level \(s_1 - s_2\); that is, it takes an object \(X\) of level \(0\) to the object \(X'\langle -s_2 \rangle[s_1]\) of level \(s_1 - s_2\).
  We can thus obtain the dimension of all morphisms of level \(\ell\) by setting \(q_1 = q\) and \(q_2 = q^{-1}\) and taking the coefficient of \(q^{\ell}\) in the expression~\eqref{eq:bigraded-homs}.
    The dimension of \(\mathbb{E}^1(X,X')\), is therefore the coefficient of~\(q\) under this specialisation.

It can be checked that under the path-length grading, the space of morphisms arising from an intersection of curves of ``crossing'' type as well as ``non pointed'' type contains exactly one morphism (up to equivalence and scalar multiplication) of level one; that is, a unique element (up to scalar multiplication) of~\(\mathbb{E}^1\).
Morphisms arising from other intersections do not contribute to~\(\mathbb{E}^1\).
This calculation proves the proposition.
\end{proof}

@@

\begin{proof}
  The proof is once again a consequence of analysing the extensions explicitly via the complexes from the Khovanov--Seidel reconstruction.
  We do not reproduce the proof here, as the argument is contained in the proof of~\cite[Thm.~5.5]{tho:06}.
  (See also~\cite[Fig.~4]{tho:06}.)
\end{proof}
Note that if we write down the relations in the Grothendieck group of our category that correspond to the extensions from each case of~\cref{prop:extensions}, then we obtain precisely the linear dependences that appear between the \(\b{g}\)-vectors in~\cref{lem:linearDependences}.
More precisely, we have that
\begin{itemize}
\item if \(\alpha\) and \(\alpha'\) are not pointed, then
  \([Y] = [X] + [X'] = [Y']\),
  from which we see \([X] + [X'] = ([Y] + [Y'])/2\);
\item if \(\alpha\) and \(\alpha'\) are crossing, then
  \([X] + [X'] = [Y] + [Y'].\)
\end{itemize}
This suggests that there should be a way to construct the \(\b{g}\)-vectors categorically in such a way that the \(\b{g}\)-vectors also map to the Grothendieck group of the category.
The \(\b{g}\)-vector of an object~\(X\) corresponding to a wiggly arc \(\alpha\) can, of course, be reconstructed from \(\alpha\).
However, at the moment we do not have a purely categorical construction of the \(\b{g}\)-vector of \(X\) that does not refer to \(\alpha\), and it would be interesting to find this.
